---
title: "EDA plan"
format: html
editor: visual
theme: 
  - lux
---

## Glimpse of the data

* First few 
* Last few 
* Random selection

```{r, echo=FALSE, message=F, warning=F}
library(tidyverse)
df <- read.csv("data/original/finalmerged_data.csv")
head(df)
tail(df)
df %>% slice_sample(n=6)
```


The idea is to quickly see if there are any immediate obvious errors

## A more detailed look

* Structure of the data - variables coded correctly? 
* Summary statistics 
* How many missing values?
```{r, echo=F}
glimpse(df)
str(df)
summary(df)
```

The variables look to be in the right format. However, I will binary code armed_conflict which makes the plotting more simple when I use this variable.

```{r, echo=FALSE}
df$armed_conflict <- factor(df$armed_conflict)
```

There are quite a few missing values for the response variables along with explanatory variables.

## Visualizations

* Response vs. explanatory variables 
* With and without oddly behaving points

```{r, echo=F}
df %>% ggplot(aes(x = maternal_mort)) +
  geom_histogram(binwidth = 10) +
  theme_classic() +
  labs(
    x = "Maternal Mortality"
  )

df %>% ggplot(aes(x = neonatal_mort)) +
  geom_histogram(binwidth = 10) +
  theme_classic() +
  labs(
    x = "Neonatal Mortality"
  )

df %>% ggplot(aes(x = under5_mort)) +
  geom_histogram(binwidth = 10) +
  theme_classic() +
  labs(
    x = "Under 5 Mortality"
  )

```

The distribution of the mortality variables are very right skewed, i.e., not normal dist.  

```{r, echo=FALSE}
ggplot(df, aes(x=armed_conflict, y=maternal_mort)) +
  geom_boxplot() +
  theme_classic()

ggplot(df, aes(x=armed_conflict, y=under5_mort)) +
  geom_boxplot() +
  theme_classic()

ggplot(df, aes(x=armed_conflict, y=neonatal_mort)) +
  geom_boxplot() +
  theme_classic()
ggplot(df, aes(x=armed_conflict, y=infant_mort)) +
  geom_boxplot() +
  theme_classic()
```

It appears that the mortality rates are higher for countries suffering from armed conflicts.

Lets see the trend over time, for now we'll take a look at Afghanistan. 

```{r, echo=FALSE}
df %>% filter(ISO == "AFG") %>% ggplot(aes(x=Year, y=maternal_mort))+
  geom_line()

ggplot(df, aes(x=Year, y=log(maternal_mort))) + 
  geom_line(aes(group=ISO)) +
  facet_wrap(~armed_conflict)

ggplot(df, aes(x=Year, y=log(infant_mort))) + 
  geom_line(aes(group=ISO))+
  facet_wrap(~armed_conflict)

ggplot(df, aes(x=Year, y=log(neonatal_mort))) + 
  geom_line(aes(group=ISO))+
  facet_wrap(~armed_conflict)

ggplot(df, aes(x=Year, y=log(under5_mort))) + 
  geom_line(aes(group=ISO))+
  facet_wrap(~armed_conflict)
```

In general, regardless of which mortality we look at, it seems to be trending downwards, which is a good sign. It's worth noting that the mortality rate seems to be fluctuating a bit for some country, however there are very few occurrences of this. In the above time-series plots, I did a log transformation on mortality, as many of the lines were clumped together.  

I also considered faceting by armed_conflict status. We see that the mortality is still trending downwards, however the mortality rate is higher for those countries that are dealing with armed conflict.

```{r}
ggplot(df, aes(x=popdens, y=maternal_mort, col=armed_conflict))+
  geom_point()
```

It does not seem that armed conflict and population density are interacting with each other to affect maternal mortality.

## Basic models

* For now, consider just dropping all the missing values

```{r}
df <- df %>% drop_na()
```


```{r, echo=F}
m1 <- lm(maternal_mort ~ armed_conflict, data = df)
summary(m1)
plot(m1)

m2 <- lm(maternal_mort ~ armed_conflict+drought+earthquake+region+GDP+OECD+OECD2023+popdens+urban+agedep+male_edu+temp, data = df)
summary(m2)
plot(m2)
```

There is a bit of a pattern in the residuals, so maybe some transformations may be helpful.  
In general, we see that armed_conflict does have a significant effect on maternal mortality in both the full and reduced model. This suggests that there could be differences in mortality due to whether a country is facing conflicts.